### 2020-08-08 thesis/chap_1.tex
fixed disc 1 from comments

Under the BAU scenarios, it can be seen that high flow values are being eroded into low flow values (and to a lesser extent, medium flow values), signifying the replacement of very high quality pixels for connectivity into very low quality pixels. We can therefore label the effects of the BAU scenarios as "directional". On the other hand, in the Reforestation scenario, high quality pixels are not being eroded (or are being matched by reforestation), but pixels of medium quality are being replaced by pixels of low quality. We can therefore label the effects of this scenario as "polarizing". Finally, we can mention that climate does not seem to noticeably change the distribution of values.\\

@@

Integrations of land use change models with climate change models are increasingly common in connectivity conservation research, but remain insufficiently frequent in the face of the threat posed by the combined effects of land use and climate change \citep{costanza_landscape_2019}. Land use change models, in particular, are lacking in biodiversity scenarios \citep{titeux_biodiversity_2016}. This absence can be explained by a range of issues, each of which the methodology and results presented in this chapter contribute to solve.

The paucity of land use change simulations, in comparison to the slew of established global climate change models available, is mentioned by \cite{costanza_landscape_2019} as a contributing factor. This paucity, we argue, can be explained by the lack of long-term land use data, but also by the opacity that surrounds land use change modelling practices. Although the former is becoming less and less of an issue with the advent of easily accessible remotely sensed data, the latter is a real obstacle to integrating land use change into connectivity models \citep{moulds_open_2015}. Here, we show a flexible framework with transparent parameterization using open-source software, which can be used to simulate landscape change in conjunction with climate change to assess connectivity or any other relevant landscape-level metric. This effort for transparency comes with a concern for taking into account uncertainty in land use change simulations, another obstacle to integration mentioned by \cite{costanza_landscape_2019}. By using MCMC (Markov Chain Monte Carlo) based simulations, we can propagate uncertainty surrounding both climate and land use change models. This is crucial, because another critique against the lack of integration is that designing conservation corridors without taking into account future land use and climate change could lead to a false sense of certainty in the efficiency of corridor designs, a mistake that could be reproduced if the uncertainty of integrated models is not better taken into account. Finally, land use change and CC are seen by the research community as impacting connectivity at different scales, with climate change models often being used in larger-scale connectivity studies than land use change. For instance, corridors designed to be adaptive to climate change often span very long distances \citep{mcguire_achieving_2016}. Yet our results show that even noticeable differences between climate scenarios can be observed, even at small scales (i.e. below the ecoregion level).

% Review results and contrast with other papers integrating LUC and CC models
Other studies have attempted to go beyond those issues and managed to integrate climate and land use change models to some degree. For example, \cite{albert_applying_2017}, on which this work is based, considered which habitat patches network would be most resilient to future change using multi scale metrics in Monteregie. \cite{leonard_landscape-scale_2017} used systematic conservation planning methods to identify core areas and linkages under an energy development land use change scenario in the northeastern US. In another study, \cite{dilts_multiscale_2016} used a range of graph theory-based metrics to forecast the changes to diverse aspects of habitat connectivity for the Mohave ground squirrel under multiple climate and land use change scenarios. Although our results, contrary to those studies, do not allow us to derive tangible predictions for our study species such as a plan for a protected area network, we still provide valuable insight for the design of such hypothetical networks.

For instance, by looking at a simple metric (mean flow), we have shown that some species will react more simply and linearly than others. We saw a simple and linear response to change for species which, in our model, had dependency on forest types that were expected to be impacted only by land use change, such as the redbacked salamander and the wood frog. This was not the case for species which are dependent on coniferous forests or mixed type forests, more at risk from climate change, and therefore more likely to suffer from the interaction of the two drivers. The results of our simple reforestation scenario highlights that it might be hard to plan for species with different needs. Restoration actions remain very uncertain: the impact of reforesting will depend on what is replanted and where it is replanted. Forest management could turn out to be critical for species dependent on forest types more vulnerable to a changing climate. A final insight is given by the width of difference between the outcome of the two scenarios for each of the species. The fact that this difference is larger for some species  than for others possibly indicates more uncertainty with regards to the real future trajectory of those species at the 2100 horizon.

Our implementation of the feature detection algorithm opens the door to more research on how future land use and climate change will impact the complexity of the species' flow. What flow complexity is and what it means for different species are open questions, and simulations such as the one presented in this chapter could represent a good opportunity to push these questions further. Although the results are imperfect in comparison to those of \cite{pelletier_forest_2017}, we still observe that the reforestation scenario manages to change some properties of the flow surfaces in such a way that we see  a rather consistent pattern in the number of features detected, possibly indicating that reforestation minimize the number of pinch points and allows for less restricted flow. More research on this feature detection technique is needed to assess its relevance in the analysis of connectivity simulation

% Limitations
This work constitutes progress over past projects on which this work is built. We covered a larger extent, improved on the land use change model from \cite{albert_applying_2017}, and demonstrated the flexibility of our framework for supporting alternative scenarios. But besides the important methodological contribution of this chapter, it is not without limitations. We see four main axes along with this work is limited and could be improved. To begin, different steps of the modelling process were limited computationally by time and computing resources, which led us to having to aggregate the raw data from a 30 meter resolution to a 90 meter resolution, leading to a loss of data and a loss in the quality of the data. A major consequence of this aggregation is that roads, which constitute an import barrier to connectivity, were not properly aggregated causing multiple arbitrary holes in the road network, severely limiting the ecological realism of the results presented in this thesis, and compromising the reliability of the flow surfaces obtained through Circuitscape. In future iteration of this work, a proper road network will need to be rasterized and added onto our resistance surfaces.

The lack of accuracy of our land use change simulations also needs to be addressed. On the one hand, this can be seen as a sign that the model did not overfit the data, while on the other hand, it is difficult to know at which point a land use change model has underfitted the data, and therefore we should not hold too much confidence in the fact that our model has been able to accurately replicate the land use change patterns observed  between 1990 and 2010. Instead, the model should be considered as an approximation and a generalization of the patterns that are observed in the region. This is possibly a good thing for planning at a regional scale, but should indicate that the results of this model should not be used for planning at smaller, local scales.

A third limitation is found in the habitat suitability analysis. The steps and parameters of the analysis were modified from \cite{rayfield_priorisation_2018} to fit our dataset and those modifications prevent us from comparing our results to their results. The analysis is also limited in the sense that multiple factors that were originally included in Rayfield's original analysis are not taken into account. For instance, the fact that black bear habitat would receive a negative multiplier if it is close to urban land is not included in our analysis (whereas it was included in Rayfield's), and should indicate that the results observed for the bear might be overestimated. Similarly, in Rayfield's analysis, the wood frog is dependent on wetlands and habitat receives a negative multiplier if it is far from wetland. Including this might change the response of this species.

Another critique could be made on the analysis of flow surfaces. Contrary to a lot of published work that wishes to capture a range of connectivity measures, by looking both at flow and at graph-based indices we have not pushed this analysis this far. \cite{rayfield_priorisation_2018} argues that there is a complementarity between outputs of Circuitscape (flow surfaces), and the outputs of graphs from habitat analyses. A more in-depth analysis of such graphs, through indices such as Probability of Connectivity for instance \citep{saura_new_2007} would be a good next step in the analysis of the results presented in this thesis.

Finally, a final limitation we want to emphasize is to be found in our reforestation scenario. This scenario has two major issues: first, it is random and overly ambitious, growing any forest patch on the landscape, and at the same rate as land is urbanized. It is very unlikely that such a scenario could be realistic. Therefore, the fact that some species are able to maintain mean flow under this scenario could be unlikely, and this scenario might represent an overly optimistic outcome, even for a scenario whose main goal is to represent an extreme of the realm of possible futures for the region.\\

@@

We successfully integrated a potential functional connectivity model for 5 umbrella species with a land use and climate change model, using circuit theory as our main method of analysis. We analyzed simple properties of the resulting flow surfaces and looked at the differences in the species response. Species who depend on forest types that are both vulnerable to land use and climate change responded in more complex, less linear, and less consistent ways than species who only depend on forest types that are mostly threatened by land use changes. While improving on related past work in the region, we open the door for other methods of flow analysis and for a more rigorous implementation of the modelling process. We hope this work can serve as a reminder that connectivity conservation planning needs to better take into account the joint effects of interacting drivers such as land use and climate change.

>>>>

Under the BAU scenarios, it can be seen that high flow values are being eroded into low flow values (and to a lesser extent, medium flow values), signifying the replacement of very high quality pixels for connectivity into very low quality pixels. We can therefore label the effects of the BAU scenarios as "directional". On the other hand, in the Reforestation scenario, high quality pixels are not being eroded (or are being matched by reforestation), but pixels of medium quality are being replaced by pixels of low quality. We can therefore label the effects of this scenario as "polarizing". Finally, we can mention that climate does not seem to noticeably change the distribution of values.

@@

The integration of land use change models with climate change models is increasingly common in connectivity conservation research. However, it remains insufficiently frequent given the threats posed by the combined effects of land use and climate change \citep{costanza_landscape_2019}. Land use change models, in particular, are often lacking in biodiversity scenarios at landscape scales \citep{titeux_biodiversity_2016}. This gap can be explained by a range of issues. The methodology and results presented in this chapter are intended to contribute a readily applicable approach .

The paucity of land use change scenarios, in comparison to the slew of established global climate change models, is mentioned by \cite{costanza_landscape_2019} as a factor contributing to this gap in the litterature. This paucity, we argue, can be explained by the lack of long-term land use data, but also by the opacity that surrounds land use change modelling practices. Although the former is becoming less and less of an issue with the advent of easily accessible remotely sensed data, the latter is a real obstacle to integrating land use change into connectivity models \citep{moulds_open_2015}. Here, we offer a flexible framework with transparent parameterization using open-source software, which can be used to simulate landscape change in conjunction with climate change to assess connectivity or any other relevant landscape-level metric.

This commitment to transparency comes with the intention to take into account uncertainty in land use change simulations, another obstacle to integration mentioned by \cite{costanza_landscape_2019}. By using MCMC (Markov Chain Monte Carlo) based simulations, we can propagate uncertainty surrounding both climate and land use change models. This is crucial, because another reason for the lack of integration is that designing conservation corridors without taking into account future land use and climate change could lead to a false sense of certainty in the efficacy of corridor designs. Finally, land use change and CC are seen by the research community as impacting connectivity at different scales, with climate change models often being used in larger-scale connectivity studies than land use change. For instance, corridors designed for climate change often span very long distances \citep{mcguire_achieving_2016}. Yet our results show that even noticeable differences between climate scenarios can be observed, even at small scales (i.e. below the ecoregion level).

% Review results and contrast with other papers integrating LUC and CC models
Other studies have attempted to go beyond those issues and managed to integrate climate and land use change models to some degree. For example, \cite{albert_applying_2017}, on which this work is based, assessed whether certain habitat networks in the Monteregie would be more resilient to future change using several multi scale metrics of connectivity. \cite{leonard_landscape-scale_2017} used systematic conservation planning methods to identify core areas and linkages under an energy development land use change scenario in the northeastern US. In another study, \cite{dilts_multiscale_2016} used a range of graph theoretic metrics to forecast the changes to diverse aspects of habitat connectivity for the Mohave ground squirrel under multiple climate and land use change scenarios. Although our results, contrary to those studies, do not allow us to derive prescriptive network design for our study species, our results do offer valuable insights for the design of such networks in the future.

For instance, by looking at a simple metric of connectivity (mean flow), we have shown that some species will react to land use change more simply and linearly than others. We saw a simple and linear response to change for species which, in our model, are dependent on forest types that were expected to be impacted only by land use change, such as the redbacked salamander and the wood frog. This was not the case for species which are dependent on coniferous forests (e.g. American marten) or mixed type forests, more at risk from climate change, and therefore more likely to suffer from the interaction between the two drivers.

The results of our simple reforestation scenario highlights that it might be hard to plan for species with different needs. Restoration action remains very uncertain: the impact of reforestation on future connectivity will depend on what is replanted and where it is done. Forest management could turn out to be critical for species dependent on forest types more vulnerable to a changing climate. A final insight is given by the width of difference between the outcome of the two scenarios for each of the species. In figures \ref{flow_linear_1} and \ref{surf_linear_1},  the end points of the most extreme response curve for warming scenarios are more spaced out in some species (short-tailed shrew and black bear) than in others (redbacked salamander, wood frog and most clearly the american marten). This pattern is consistent for both mean flow and number of features detected (complexity of the flow). The fact that this difference is larger for some species  than for others possibly indicates more uncertainty with regards to the real future trajectory of those species at the 2100 horizon. It could signify that restoration action is likely to bring a better “return on investment” for species that have a larger gap between outcomes than for other species whose gap suggests their trajectory is more determined.

Our implementation of the feature detection algorithm opens the door to more research on how future land use and climate change will impact the complexity of the species' flow. What flow complexity is and what it means for different species are open questions, and simulations such as the one presented in this chapter are a good opportunity to push these questions further. To our knowledge, the only previous published work to have used this method is \cite{pelletier_forest_2017}. Their implementation of the SURF algorithm allowed them to quantify the number of pinch-points of different scales. \citeauthor{pelletier_forest_2017} applied this method to a much larger extent than we did, and comparing the results would be difficult. However, a few observations are worth mentioning. First, their results produced clean outputs, with clean linear pinch points. Our results are less clean, with more globular features being detected at the same time. We argue this is due to the fact that \citeauthor{pelletier_forest_2017} had less variance in their resistance values than us, as they used a simpler habitat suitability measure. Furthermore, \citeauthor{pelletier_forest_2017} did not consider species-specific responses, and our different species might require different SURF parameterizations (threshold values, number of octaves). Second, although the results are imperfect in comparison to those of \cite{pelletier_forest_2017}, we observed that reforestation scenario changed some properties of the flow surfaces in such a way that we see  a rather consistent pattern in the number of features detected, possibly indicating that reforestation minimizes the number of pinch points and allows for less restricted flow. More research on this feature detection technique is needed to assess its value in the analysis of connectivity simulations \cite{pelletier_forest_2017}.

% Limitations
This work builds on past projects in this region. We covered a larger spatial extent, improved the land use change model started by \cite{albert_applying_2017}, and demonstrated the flexibility of our framework for supporting alternative scenarios. But besides the important methodological contribution of this chapter, it is not without limitations. We see four main axes along with this work can be improved. To begin, different steps of the modelling process were limited computationally by time and computing resources, which led us to having to aggregate the raw data from a 30 meter resolution to a 90 meter resolution, leading to a loss of information. A major consequence of this aggregation is that roads, which constitute an important barrier to connectivity, would have been broken up by the aggregation process. Instead, they had to be aggregated out of the raster surface and a road network was rasterized and overimposed onto the newly aggregated raster. This led us to keep only major roads, an outcome that means we ignored the impact of smaller (<90m) roads on connectivity. One possible solution would be to aggregate all roads but to weight the resistance differently based on road type.

Our historical land use change simulations did not accurately capture land use change between 1990-2010. On the one hand, this can be seen as a sign that the model did not overfit the data, while on the other hand, it is difficult to know at which point a land use change model has underfitted the data. However, we should not hold too much confidence in the fact that our model has been able to accurately replicate the land use change patterns observed  between 1990 and 2010. Instead, the model should be considered as an approximate and generalization simulator of the land use patterns that are observed in the region. This is possibly a good thing for planning at a regional scale, but should indicate that the results of this model should not, at this stage, be used for planning at smaller, local scales.

A third limitation is found in the habitat suitability analysis. The steps and parameters of the analysis were modified from \cite{rayfield_priorisation_2018} to fit our dataset and those modifications prevent us from comparing our results to their results. The analysis is also limited in the sense that multiple factors that were originally included in Rayfield's original analysis are not taken into account. For instance, the fact that black bear habitat would receive a negative multiplier if it is close to urban land is not included in our analysis (whereas it was included in Rayfield's), and should indicate that the results observed for the bear might be overestimated. Similarly, in Rayfield's analysis, the wood frog is dependent on wetlands and habitat receives a negative multiplier if it is far from wetland. Including this might change the response  for this species.

Another critique could be made on the analysis of flow surfaces. A more in-depth analysis of such graphs, through indices such as Probability of Connectivity for instance \citep{saura_new_2007} would be a good next step in the analysis of the results presented in this thesis. Future work could also combine flow and graph-based indices of connectivity. \cite{rayfield_priorisation_2018} argues that there is a complementarity between outputs of Circuitscape (flow surfaces), and the outputs of graphs from habitat analyses which provide a more complete picture of the effects of land use and climate change.

Finally, some limitations of our reforestation scenario are worth mentioning. This scenario made two assumptions: first, it is random and overly optimistic, because it grows any forest patch on the landscape, and at the same rate as land is urbanized. This scenario was not meant to be realistic but it provides an ideal case against which more realistic scenarios can be compared. Future work is needed to develop a range of scenarios of regional land use and climate change for this study region. Ongoing work with stakeholders (e.g. municipal and provincial government and conservation NGOs) will ensure that these are developed. When these scenarios are available, the methods presented in this chapter will prove useful for exploring their impacts on the habitat connectivity of the region’s fauna \\

@@

We successfully integrated a potential functional connectivity model for 5 umbrella species with a land use and climate change model. We used circuit theory as our main method of analysis. We analyzed simple properties of the resulting flow surfaces and looked at the differences in the species’ responses. Species that depend on forest types that are both vulnerable to land use and climate change responded in more complex, less linear, and less consistent ways than species that only depend on forest types that are mostly threatened by land use changes. While improving on recent work in the region, our approach also opens the door to other methods of flow analysis and to a more rigorous and open implementation of the modelling process. We hope this work can support future connectivity conservation planning that needs to take into account the joint and interacting effects of land use and climate change. \\
